### 2023-04-24 response-to-audit-report.md

> **File `specification-v3-rc1.pdf`, Section `3.6.3 Contract Does Not Hold Funds After it Closes`, page *32***
>
> The statement in natural language looks unconnected from the proposed formula. Otherwise, it is unclear how not holding funds forever is a consequence of producing no warnings.

The statement was indeed unconnected from the description. In [PR-186](https://github.com/input-output-hk/marlowe/pull/186) we refactored the guarantees section and now the section "3.2 Finite contracts"
includes the theorem `timedOutTransaction_closes_contract` which shows that after closing a timeouted contract, the accounts are emptied.

@@

### 2.2.26 Different format for lemma statement *(Severity: Low)*

> **File `specification-v3-rc1.pdf`, Sections `3.6.2 All Contracts Have a Maximum Time`, page *32***
>
> The lemma is stated using the proof derivation tree format as opposed to the rest of the specification and the Isabelle code.

This was addressed in [PR-168](https://github.com/input-output-hk/marlowe/pull/168)

@@

### 2.2.27 Function `isClosedAndEmpty` is unexplained *(Severity: Low)*

> **File `specification-v3-rc1.pdf`, Section `3.6.2 All Contracts Have a Maximum Time`, page *32***
>
> The function `isClosedAndEmpty` needs to be explained.

In [PR-186](https://github.com/input-output-hk/marlowe/pull/186) we add a small explanation, in the future we'll probably expand on this by refactoring the guarantees as literate programming.

@@

### 2.2.28 Top-down definitions *(Severity: Low)*

> **File `specification-v3-rc1.pdf`, Section `2`**
>
> In Section 2, the order of definitions is reversed, and the reader is thus faced with functions which call other functions that have not been introduced yet, despite the claim in Section 1.3 that the definitions will be presented bottom-up.

This section was redone in [PR-182](https://github.com/input-output-hk/marlowe/pull/182)

@@

### 2.2.29 No mention of Isabelle lemmas in specification *(Severity: Low)*

> **File `specification-v3-rc1.pdf`, Multiple sections ``**
>
> Generally, readability can be improved by mentioning the Isabelle lemma names alongside their statements. This way, it would be much easier to search for the actual Isabelle code and proofs matching the informal specification text, and compare the two.

In [PR-186](https://github.com/input-output-hk/marlowe/pull/186) we added the names of the theories in the PDF.

@@

## 2.3 Lemmas and proofs

### 2.3.1 Unnecessarily large proofs *(Severity: Medium)*

> **Several Isabelle files, several lemmas**
>
> Some Isabelle proofs are written with long apply-scripts, where Isar would document the proof better. Proofs could also be split using more auxiliary lemmas.
>
> As the proofs stand, it is hard to figure out why a proof step fails, after changes elsewhere required a proof to be updated. Since the newly-failing proof step was designed with specific goals in mind, and changes in the code may lead to it facing a different set of goals, the maintainer might need to reconstruct the whole structure of the proof from an older version to infer state that Isabelle produces at each step.
>
> What Isar brings is making the intention of the author explicit at every step of the proof. This helps the maintainer of the proofs and fixes the concerns mentioned above.
>
>  will likely have to update the proofs. We conjecture that it will happen at least every time they target a new platform. In the case of Cardano, they need to extend the semantics to explain Merkleization. Another action that would make long proofs easier to understand is to split them using more auxiliary lemmas, thus feeding the information to the reader in smaller chunks.
>
> Some examples of large proofs:
> -   in `MoneyPreservation.thy`, lemmas `reduceContractStep_preserves_money` and\ `reductionLoop_preserves_money`
> -   in `SingleInputTransactions.thy`, lemmas `applyAllInputsPrefix_aux`,\ `computeTransactionIterative`, and `computeTransactionStepEquivalence_error`

We agree with this observation. New lemmas and theories are proven with this recommendation in mind and some theories were refactored (like the Asset preservation). It will take some time until all theories are in this format.

@@

Same response as 2.3.1

### 2.3.5 Involved proof of `insert_valid` *(Severity: Low)*

> **File `MList.thy`, theorem `insert_valid`, line *66***
>
> The proof of `insert_valid` sprouts three other lemmas of difficult characterization: `insert_valid_aux`, `insert_valid_aux2`, and `insert_valid_aux3`. These lemmas make assumptions with implications that get in the way of understanding them in isolation.
>
> An alternative to make the proof pieces more reusable is to use instead the following set of lemmas, which also offers insight on how function `insert` interacts with predicates `sorted` and `distinct`:
>
> ``` {.text escapeinside="||" mathescape="true"}
>   lemma insert_set:
>     "set (map fst (insert a b xs)) = set (map fst xs) |$\cup$| { a }"
>
>   lemma insert_sorted:
>     "List.sorted (map fst c) |$\Longrightarrow$| List.sorted (map fst (insert a b c))"
>
>   lemma insert_distinct :
>     "List.distinct (map fst c)
>      |$\Longrightarrow$|
>      List.sorted (map fst c)
>      |$\Longrightarrow$|
>      List.distinct (map fst (MList.insert a b c))"
> ```
>
> which then can be combined in the proof of `insert_valid` as follows:
>
> ``` {.text escapeinside="||" mathescape="true"}
>   theorem insert_valid2 : "valid_map c |$\Longrightarrow$| valid_map (MList.insert a b c)"
>     using insert_sorted[of c a b] insert_distinct[of c a b] by fastforce
> ```
>
> The proofs of the lemmas can be found in […].

We agree that the proposed prove is easier, but the lemma will become obsolete once we replace the custom MList with
`HOL-Library/finiteMap`, which addresses other problems found in the report.

@@

### 2.3.6 Repeated verbose expression *(Severity: Low)*

> **File `MoneyPreservation.thy`, lemma `removeMoneyFromAccount_preservation`, line *202***
>
> The expression
>
> ``` text
>   giveMoney
>     accId
>     (Party p)
>     tok
>     paidMoney
>     (updateMoneyInAccount accId tok (balance - paidMoney) accs)
> ```
>
> is large and used in other lemmas as well. It would need to be moved to a separate function to save the effort of reading it repeteadly.

The MoneyPreservation was refactored into Asset preservation and now there is a single reference to `giveMoney`. As mentioned in observation 2.2.11, giveMoney is likely to be removed in the future.

@@

### 2.3.7 Inconsistent variable name `valTrans` *(Severity: Low)*

> **File `MoneyPreservation.thy`, lemma `transferMoneyBetweenAccounts_preserves_aux`, line *257***
>
> The lemma uses a variable `valTrans` where other proofs use the name `paidMoney`. To convey the meaning of the variable faster, the same name should be used consistently in all places.

This theory was refactored in [PR-161](https://github.com/input-output-hk/marlowe/pull/161) and the lemma was no longer needed.

@@

This theory was refactored in [PR-161](https://github.com/input-output-hk/marlowe/pull/161) and the lemma was no longer needed.

### 2.3.11 Missing theorem regarding `playTrace` *(Severity: Low)*

> **File `PositiveAccounts.thy`, `playTrace preserves valid and positive state`**
>
> There is no theorem that `playTrace` keeps the state valid and positive when given a state which is valid and positive. This trivially follows from `playTraceAux_preserves_validAndPositive_state` but no such theorem is present.

The valid and positive preservation is needed in lower-level functions so that higher-level functions can use that fact when proving other properties. This makes `playTrace` not needing that proof as it is the highest-level function. The theorem  could be useful on its own and might get added in the future.

@@

### 2.3.12 Unconcise goal in `reduceContractStepPayIsQuiescent` *(Severity: Low)*

> **File `QuiescentResult.thy`, lemma `reduceContractStepPayIsQuiescent`, line *8***
>
> This lemma does not express its goal concisely, as it makes no mention of `reduceContractStep` in the formulation. Changing the first assumption to $\texttt{reduceContractStep}\ \mathit{env}\ \mathit{sta}\ (\texttt{Pay}\ \mathit{x21}\ \mathit{x22}\ \mathit{tok}\ \mathit{x23}\ \mathit{x24})$ makes more explicit in which contexts this lemma can be useful. Modifying this assumption requires an additional `apply simp` to be added to the proof (before line 30) for the lemma to go through. Further, an additional `apply simp` will need to be added in lemmas `reduceContractStepIsQuiescent` (before line 44) and `timedOutReduce_only_quiescent_in_close` (`Timeout.thy`, before line 128) as well.

We agree on this observation and it will addressed in a future refactor of the theory file.

@@

### 2.3.13 Misleading lemma names *(Severity: Low)*

> **File `PositiveAccounts.thy`, lemma `reduceOne_gtZero`, line *80***
>
> This lemma should be renamed as `refundOne_gtZero`.
>
> **File `QuiescentResult.thy`, lemma `reduceOneIsSomeIfNotEmptyAndPositive`, line *32***
>
> This lemma should be renamed as `refundOneIsSomeIfNotEmptyAndPositive`.
>
> **File `TransactionBound.thy`, lemma `computeTransaction_decreases_maxTransaction_aux`, line *240***
>
> This lemma should be renamed as `applyAllInputs_decreases_maxTransactions` or `applyAllInputs_reduced_decreases_maxTransactions`.

We agree on these observations and they will likely be addressed when we refactor the corresponding files.

@@

We agree on this observation and it will addressed in a future refactor of the theory file.

### 2.3.16 Unused and undocumented lemmas *(Severity: Low)*

> **Several Isabelle files, several lemmas**
>
> Some lemmas are never used, and they would need comments motivating their presence:
> 1.  In file `MoneyPreservation.thy`, line 257, lemma `transferMoneyBetweenAccounts_preserves_aux`.
> 2.  In file `QuiescentResult.thy`
>     1.  Line 5, lemma `reduceOne_onlyIfNonEmptyState`
>     2.  Line 153, lemma `reduceContractUntilQuiescent_ifDifferentReduced`
> 3.  In file `PositiveAccounts.thy`, line 66, lemma `positiveMoneyInAccountOrNoAccount_sublist_gtZero`. Furthermore, it is identical to `positiveMoneyInAccountOrNoAccount_gtZero_preservation`, but with an additional assumption `money > 0`.
> 4.  In file `ValidState.thy`
>     1.  Line 9, lemma `valid_state_valid_choices`
>     2.  Line 13, lemma `valid_state_valid_valueBounds`
> 5.  In file `SingleInputTransactions.thy`, line 1214, lemma `traceListToSingleInput_isSingleInput`. It is mentioned in a commented out line in `StaticAnalysis.thy`. Furthermore, the lemma can be expressed more concisely as `$$\llparenthesis \mathit{interval} = \mathit{inte}, \mathit{inputs} = \mathit{inp\_h}\ \#\ \mathit{inp\_t} \rrparenthesis\ \#\ t = \mathit{traceListToSingleInput}\ \mathit{t2} % \mathrel{% \mbox{\fontfamily{cmr}\fontencoding{OT1}\selectfont=}}% \joinrel\Rightarrow\mathit{inp\_t} = []$$`

* Lemma `transferMoneyBetweenAccounts_preserves_aux` was removed in [PR-161](https://github.com/input-output-hk/marlowe/pull/161)
* Lemma `reduceOne_onlyIfNonEmptyState`, `valid_state_valid_choices` were removed in [PR-168](https://github.com/input-output-hk/marlowe/pull/168)
* Lemma `reduceContractUntilQuiescent_ifDifferentReduced` was promoted to a theorem with an explanation in  [PR-168](https://github.com/input-output-hk/marlowe/pull/168)
* Lemma `positiveMoneyInAccountOrNoAccount_gtZero_preservation` was removed in favour of `positiveMoneyInAccountOrNoAccount_gtZero_preservation` in [PR-168](https://github.com/input-output-hk/marlowe/pull/168)

@@

### 2.3.17 Redundant `reduceContractStep` lemmas *(Severity: Low)*

> **File `MoneyPreservation.thy`, lemma `reduceContractStep_preserves_money_acc_to_acc_aux`, line *310***
>
> This lemma is weaker than `transferMoneyBetweenAccounts_preserves`. If we replace its usage at line 351 with `transferMoneyBetweenAccounts_preserves`, the proof goes through.

This lemma was removed in [PR-161](https://github.com/input-output-hk/marlowe/pull/161)

@@

This lemma was removed in [PR-161](https://github.com/input-output-hk/marlowe/pull/161)

### 2.3.19 Duplicated lemmas *(Severity: Low)*

> **File `PositiveAccounts.thy`, theorems `computeTransaction_gtZero, accountsArePositive`, lines *257, 369***
>
> These theorems are identical (modulo variable names), and one of them should be removed.
>
> **File `PositiveAccounts.thy, ValidState.thy`, lemma `valid_state_valid_accounts`, lines *381, 5***
>
> This lemma is defined twice, once in each of these files. One of them should be removed.

Lemma `accountsArePositive` was removed in favour of `computeTransaction_gtZero` in [PR-168](https://github.com/input-output-hk/marlowe/pull/168)

@@

### 2.3.20 Redundant `computeTransaction` lemmas *(Severity: Low)*

> **File `ValidState.thy`, lemmas `computeTransaction_preserves_valid_state_aux, computeTransaction_preserve_valid_state`, lines *160, 176***
>
> If `computeTransaction_preserves_valid_state_aux` is rewritten to have the same formulation as `computeTransaction_preserves_valid_state`, then the lemma (with the exact same proof) is still accepted, and these lemmas become duplicates of each other. Thus, no auxiliary lemma is needed.

Lemma `computeTransaction_preserves_valid_state_aux` was removed and `computeTransaction_preserve_valid_state` simplified in [PR-168](https://github.com/input-output-hk/marlowe/pull/168)

@@

These issues were resolved in [PR-168](https://github.com/input-output-hk/marlowe/pull/168) and [PR-169](https://github.com/input-output-hk/marlowe/pull/169)

@@

### 2.3.23 Inconsistent style when applying constructor *(Severity: Low)*

> **File `SingleInputTransactions.thy`, lemmas `beforeApplyAllLoopIsUseless, fixIntervalOnlySummary`, lines *275, 398***
>
> The lines mentioned in these lemmas display the resulting constructor before the function application, which differs from the general style in the rest of the codebase.

These observation regarding readability will be addressed when we refactor the theory.

@@

These observation regarding readability will be addressed when we refactor the theory.

### 2.3.25 Typo with "independet" in multiple lemmas *(Severity: Low)*

> **File `SingleInputTransactions.thy`, lemmas `applyAllLoop_independet_of_acc_error1, applyAllLoop_independet_of_acc_error2`, lines *977, 987***
>
> In both of these lemmas, there is a typo with the word "independet".

This issue was resolved in [PR-168](https://github.com/input-output-hk/marlowe/pull/168)

@@

### 2.3.26 Poorly named `acc` lemmas *(Severity: Low)*

> **File `SingleInputTransactions.thy`, lemmas `applyAllLoop_independet_of_acc_error1, applyAllLoop_independet_of_acc_error2`, lines *977, 987***
>
> It is unclear what `acc` refers to in these lemma names, as the lemmas are about the independence of warnings and payments, not accounts.

This observation regarding readability will be addressed when we refactor the theory.

@@

This observation regarding readability will be addressed when we refactor the theory.

### 2.3.28 `playTrace_only_accepts_maxTransactionsInitialState` not written as `theorem` *(Severity: Low)*

> **File `TransactionBound.thy`, lemma `playTrace_only_accepts_maxTransactionsInitialState`, line *316***
>
> This lemma seems like the main result of this file. Assuming it is an important result, we recommend writing it as a `theorem` rather than a `lemma`.

We agree on the observation, will likely promote to theorem in the future

@@

We agree on this observation and it will addressed in a future refactor of the theory file.

### 2.3.32 `playTrace_preserves_valid_state` not written as `theorem` *(Severity: Low)*

> **File `ValidState.thy`, lemma `playTrace_preserves_valid_state`, line *194***
>
> This lemma seems like the main result of this file. Assuming it is an important result, we recommend writing it as a `theorem` instead.

The lemma was promoted to a theorem in [PR-168](https://github.com/input-output-hk/marlowe/pull/168)

@@

We removed the unnecessary assumptions in [PR-168](https://github.com/input-output-hk/marlowe/pull/168)

@@

## 2.4 Isabelle implementation

### 2.4.1 Variable shadowing in `applyAllLoop` *(Severity: Medium)*

> **File `Semantics.thy`, function `applyAllLoop`, line *575***
>
> The $\texttt{cont}$ variable introduced by the pattern match shadows another $\texttt{cont}$ variable, coming from the pattern match of an outer case expression, making the function harder to follow while also making it more error-prone to future changes.

The variable was renamed in [PR-168](https://github.com/input-output-hk/marlowe/pull/168)

@@

### 2.4.2 Undescriptive name `moneyInPayment` *(Severity: Low)*

> **File `MoneyPreservation.thy`, function `moneyInPayment`, line *5***
>
> The name of the function can be more precise. Perhaps `moneyInPaymentToParty` or `moneyInExternalPayment` would work.

The theory was refactored into AssetsPreservation and the function is now called `assetsInExternalPayment`

@@

### 2.4.3 Typo in section name *(Severity: Low)*

> **File `OptBoundTimeInterval.thy`, line `37`**
>
> Typo in section name: "Interval intesection".

The typo was fixed in [PR-168](https://github.com/input-output-hk/marlowe/pull/168)

@@

The typo was fixed in [PR-168](https://github.com/input-output-hk/marlowe/pull/168)

### 2.4.5 Unclear need for multiple formulations for positive accounts *(Severity: Low)*

> **File `PositiveAccounts.thy`, `Throughout`**
>
> It is unclear what the use is for multiple formulations (and lemmas about) positive accounts. The first formulation (with the theorems `playTraceAux_gtZero` and `playTrace_gtZero`) is not used in any other modules but the alternative formulation is used instead. If both formulations are relevant, then it should be explained why.

We agree that it is confusing and will be refactored in the future.

@@

### 2.4.6 Variable name discrepancy in `reductionLoop` *(Severity: Low)*

> **File `Semantics.thy`, function `reductionLoop`**
>
> When comparing this function against `specification-v3-rc1.pdf`, different names are used for a let-bound variable. It is `a` in the pdf and `newPayments` in the file `Semantics.thy`. There are similar issues in the function `reduceContractStep` in the equation for the `If` case, and in the function `giveMoney`.

This was an artifact of showing the exported haskell code. It is resolved in [PR-182](https://github.com/input-output-hk/marlowe/pull/182)

@@

The typo was fixed in [PR-168](https://github.com/input-output-hk/marlowe/pull/168)

### 2023-04-24 response-to-audit-report.md

### 2.4.8 Unclear function name `calculateNonAmbiguousInterval` *(Severity: Low)*

> **File `Semantics.thy`, function `calculateNonAmbiguousInterval`, line *725***
>
> The meaning of the function is not obvious. It needs a comment to explain it.

This function was documented in [PR-168](https://github.com/input-output-hk/marlowe/pull/168)
and moved to `TimeRange.thy` in [PR-182](https://github.com/input-output-hk/marlowe/pull/182). The function is not part of the Marlowe spec yet, it is a helper we hope will be useful in the future to construct the transactions.

@@

### 2.4.9 Non-modularized file `SingleInputTransactions.thy` *(Severity: Low)*

> **File `SingleInputTransactions.thy`, `Splitting File`**
>
> This file is very long, and it covers more than just single-input transactions. For instance, about 530 lines at the beginning are rather dedicated to idempotence of certain operations. Then, the lemmas around lines 530 -- 700 focus on "well-foundedness" of the recursion on contract steps. Then there is also a clear block of lemmas about "distributivity" of semantics over transaction lists.
>
> Splitting the module, grouping the related lemmas, would help understanding the relationships between the groups.

We agree with the observation and it will be addressed in a future refactor.

@@

### 2.4.10 Misleading function names *(Severity: Low)*

> **File `SingleInputTransactions.thy`, function `inputsToTransactions`, line *9***
>
> This function name is not very descriptive of its meaning. It takes a transaction (both a time interval and a list of inputs) and returns a list of transactions at the same interval containing a single input each. A name like `splitTransactionIntoSingleInputTransactions` would convey better what the input and the output are.
>
> Moreover, the code would be cleaner if the function takes a single argument of type `Transaction`, instead of asking the caller to rip apart its fields.
>
> **File `SingleInputTransactions.thy`, function `traceListToSingleInput`, line *18***
>
> This function name is not descriptive of what it does. Perhaps a more telling name could be `splitTransactionsIntoSingleInputTransactions`.
>
> **File `SingleInputTransactions.thy`, function `isSingleInput`, line *1222***
>
> This function should be renamed or repurposed. If renamed, `allAreSingleInput` more accurately reflects the meaning of the function. If repurposed, it should check that a single transaction has a single input, and `all isSingleInput` can be used to express the current behavior.

We agree on the observations, for the moment some simple refactor and comments were added as part of
[PR-180](https://github.com/input-output-hk/marlowe/pull/180/files#diff-638dc578b752590b5f2bace45f1e1ae9e522351f62056e6f92aa272424657b03). Further refactor and optional renaming might happen in the future.

@@

We agree on this observation and it will addressed in a future refactor of the theory file.

### 2.4.12 Duplicated `isValidInterval` function *(Severity: Low)*

> **File `TimeRange.thy`, function `isValidInterval`, line *231***
>
> This function duplicates $\texttt{validTimeInterval}$ from $\texttt{OptBoundTimeInterval.thy}$, and the latter has certain additional properties proven about it specifically, so it makes sense to use the latter in both cases.

The function `isValidInterval` was removed in favour of `validTimeInterval` in [PR-168](https://github.com/input-output-hk/marlowe/pull/168/files#diff-9c936a78b5d8eed5683713bb31c59366b2a0794c5f11e9807f7fa61d1b11b7c2L230)

@@

> **Helper function `evalValue, evalObservation`, lines *391 (Semantics.hs), 34 (Semantics.thy)***
>
> `evalValue` and `evalObservation` differ from the Isabelle implementation in the introduction of auxiliary variables to abbreviate the recursive calls. The comparison would be simpler if both definitions were consolidated.

The Plutus implementation differs slightly from the Isabelle for efficiency and to conform to Plutus's different restrictions on recursive calls.

> **Helper function `evalValue`, lines *395 (Semantics.hs), 35 (Semantics.thy)***
>
> The Isabelle implementation should use the helper function `moneyInAccount` instead of inlining its definition, so as to maintain consistency with the Haskell implementation.

This was addressed in [PR-182](https://github.com/input-output-hk/marlowe/pull/182)

@@

This misstatement is remedied in commit [8691f335](https://github.com/input-output-hk/marlowe-cardano/commit/8691f3351e55758cc750eee0800b1458f0a8c554).

### 2.6.15 Timeout boundary differs from the specification *(Severity: Low)*

> **File `Semantics.hs`, type `reduceContractStep`, line *518***
>
> The specification mentions
>
> > If a valid Transaction is computed with a TimeInterval with a start time bigger than the Timeout t, the contingency continuation c is evaluated.
>
> where "bigger" implies strict inequality, while the code makes non-strict comparison. This difference needs to be acknowledged and further explained in the specification.

This misstatement is remedied in [PR-182](https://github.com/input-output-hk/marlowe/pull/182)
